% ACL 2026 / EMNLP 2026 Format
\documentclass[11pt]{article}
\usepackage{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{subcaption}

\title{A Comprehensive Benchmark of Retrieval Methods for\\Text-and-Table RAG Systems}

\author{
  Meftun Akarsu \\
  Technische Hochschule Ingolstadt \\
  \texttt{meo5963@thi.de} \\
  \And
  Christopher Mierbach \\
  Radiate \\
  \texttt{christopher@radiaite.com} \\
}

\begin{document}
\maketitle

\begin{abstract}
Retrieval-Augmented Generation (RAG) systems critically depend on retrieval quality,
yet no comprehensive comparison of modern retrieval methods exists for documents
containing both text and tabular data. We systematically benchmark 15+ retrieval
strategies---spanning sparse (BM25, SPLADE), dense (multiple embedding models),
late interaction (ColBERT), hybrid fusion (RRF, convex combination), query expansion
(HyDE, multi-query), index augmentation (HyPE, contextual retrieval), and
hierarchical approaches (parent-child, RAPTOR)---on the T\textsuperscript{2}-RAGBench
dataset comprising 23,088 financial QA pairs over 7,318 documents. We evaluate both
retrieval quality (Recall@k, MRR, nDCG) and end-to-end generation quality (Number Match,
F1, BERTScore) with statistical significance testing. Our results show that
[KEY FINDING 1], [KEY FINDING 2], and [KEY FINDING 3]. We provide practical
recommendations for RAG system design and release our full benchmark code.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

Retrieval-Augmented Generation (RAG) has emerged as the dominant paradigm for
grounding large language models in external knowledge \cite{lewis2020rag}.
The quality of the retrieval component is critical: if relevant documents are not
retrieved, no amount of generation capability can compensate.

While numerous retrieval methods have been proposed---from classical sparse
retrieval (BM25) to learned dense representations, late interaction models,
and hybrid approaches---practitioners face a fragmented landscape with limited
guidance on which method to use, especially for \textbf{heterogeneous documents
containing both text and tables}.

Financial documents exemplify this challenge: earnings reports, SEC filings,
and annual reports contain narrative text interspersed with numerical tables.
Retrieving the correct context for a question like ``What was the year-over-year
revenue growth?'' requires matching both textual descriptions and specific
numerical cells---a task that strains purely semantic retrieval methods.

The recently introduced T\textsuperscript{2}-RAGBench \cite{strich2026t2ragbench}
provides a rigorous benchmark for this setting, with 23,088 question-context-answer
triples from 7,318 financial documents. However, the original paper evaluated
only six retrieval methods with two metrics. Many modern approaches---ColBERT,
contextual retrieval, HyPE, RAPTOR, and recent reranking models---remain untested
on this benchmark.

\paragraph{Contributions.} We make the following contributions:
\begin{enumerate}
  \item We present the most comprehensive RAG retrieval benchmark to date,
    systematically evaluating 15+ methods on T\textsuperscript{2}-RAGBench.
  \item We provide multi-dimensional evaluation covering retrieval metrics
    (Recall@k, MRR, nDCG, MAP), generation metrics (Number Match, F1, BERTScore),
    and RAG-specific metrics, with statistical significance testing.
  \item We conduct extensive ablation studies isolating the effects of embedding
    models, chunk sizes, rerankers, and fusion strategies.
  \item We provide actionable recommendations with cost-accuracy trade-off
    analysis for practitioners building text-and-table RAG systems.
\end{enumerate}

% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{RAG Retrieval Methods.}
Retrieval methods for RAG span a broad spectrum of representation strategies.
Classical sparse retrievers such as BM25 remain surprisingly competitive baselines,
particularly in zero-shot out-of-domain settings on the BEIR benchmark \citep{thakur2021beir},
owing to their lexical precision, interpretability, and scalability.
Learned sparse models like SPLADE \citep{formal2021splade} extend this paradigm by
learning neural term expansion weights, with recent work showing that decoder-only
LLM backbones further improve sparse retrieval quality \citep{doshi2024mistralsplade}.
On the dense side, dual-encoder architectures pioneered by DPR \citep{karpukhin2020dpr}
encode queries and passages into a shared embedding space for efficient approximate nearest
neighbor search. Subsequent embedding models---E5 \citep{wang2022e5}, E5-Mistral
\citep{wang2024e5mistral}, and BGE-M3 \citep{chen2024bgem3}---have progressively
advanced the state of the art on MTEB \citep{muennighoff2023mteb} and its multilingual
successor MMTEB \citep{enevoldsen2025mmteb} through contrastive pre-training at scale
and instruction-tuned embedding generation.
Late-interaction models such as ColBERTv2 \citep{santhanam2022colbertv2}
and its multilingual extension Jina-ColBERT-v2 \citep{sturua2024jinacolbert}
offer a middle ground: they retain per-token representations for fine-grained matching
while enabling pre-computation of document representations.
Hybrid retrieval combines sparse and dense signals via Reciprocal Rank Fusion
\citep{cormack2009rrf} or convex score combination, consistently improving recall
by 15--30\% over single-method pipelines \citep{li2025ragbestpractices}.
Two-stage reranking with cross-encoder models \citep{sun2023rankgpt} further refines
ranked lists, with recent benchmarks reporting up to 28\% nDCG@10 improvement at modest
latency cost \citep{gao2024ragsurvey}.

\paragraph{Advanced Retrieval Strategies.}
Beyond standard retrieve-then-generate pipelines, several strategies manipulate
queries or indices to improve retrieval quality.
\emph{Query expansion} approaches include HyDE \citep{gao2022hyde}, which generates a
hypothetical answer document at query time and retrieves using its embedding, and
RAG-Fusion \citep{raudaschl2023ragfusion}, which issues multiple LLM-rewritten query
variants and merges results via RRF.
In contrast, \emph{index augmentation} methods enrich document representations at
indexing time: Contextual Retrieval \citep{anthropic2024contextual} prepends
LLM-generated chunk summaries to each passage, while HyPE \citep{vake2025hype}
precomputes hypothetical questions per chunk, transforming retrieval into
question--question matching and improving context precision by up to 42 percentage
points without added query-time latency.
Late Chunking \citep{guenther2024latechunking} preserves cross-chunk context by
applying long-context embedding models before splitting.
Hierarchical retrieval strategies include RAPTOR \citep{sarthi2024raptor}, which
builds recursive tree-structured summaries, and GraphRAG \citep{edge2024graphrag},
which constructs entity-relation graphs for query-focused summarization.
On the adaptive retrieval front, Self-RAG \citep{asai2023selfrag} trains models
to self-reflect on whether retrieval is needed and to critique generated outputs,
while CRAG \citep{yan2024crag} evaluates retrieved document quality and triggers
corrective web searches when confidence is low.
More recently, Agentic RAG architectures embed autonomous agents that dynamically
plan multi-step retrieval, select among heterogeneous tools, and iteratively refine
context \citep{singh2025agenticrag}.

\paragraph{Text-and-Table Question Answering.}
Answering questions over documents containing both text and tables poses unique
challenges: models must locate relevant evidence across heterogeneous modalities
and often perform numerical reasoning.
\citet{chen2020hybridqa} introduced HybridQA, the first large-scale dataset requiring
multi-hop reasoning over linked Wikipedia tables and passages.
\citet{chen2021ottqa} extended this to an open-domain setting in OTT-QA, where models
retrieve from over 400K table and 5M passage candidates.
In the financial domain, FinQA \citep{chen2021finqa} provides expert-annotated
question--program pairs over earnings reports, while TAT-QA \citep{zhu2021tatqa}
focuses on diverse numerical operations (addition, comparison, counting) over
hybrid tabular-textual contexts from financial filings.
ConvFinQA \citep{chen2022convfinqa} extends FinQA to multi-turn conversational
reasoning.
A recent comprehensive survey of table QA in the LLM era \citep{nan2025tableqasurvey}
highlights that despite advances in LLM-based table understanding, retrieval of the
correct heterogeneous context remains a primary bottleneck---a finding that directly
motivates our retrieval-focused evaluation on T\textsuperscript{2}-RAGBench
\citep{strich2026t2ragbench}, which unifies FinQA, ConvFinQA, and TAT-DQA subsets
into a single retrieval benchmark with 23,088 queries over 7,318 financial documents.

\paragraph{RAG Benchmarks and Evaluation.}
Standardized benchmarks have been critical for measuring retrieval and generation
progress.
BEIR \citep{thakur2021beir} established zero-shot evaluation across 18 diverse
retrieval datasets, while MTEB \citep{muennighoff2023mteb} and MMTEB
\citep{enevoldsen2025mmteb} broadened scope to 8 embedding tasks across 500+
multilingual datasets.
For knowledge-intensive NLP, KILT \citep{petroni2021kilt} unified five task types
over a shared Wikipedia snapshot.
RAG-specific benchmarks include RGB \citep{chen2024rgb}, which tests noise robustness,
negative rejection, and counterfactual robustness; CRAG \citep{yang2024crag},
which spans five domains with temporal dynamics; and RAGBench \citep{friel2024ragbench},
which provides 100K examples with the explainable TRACe evaluation framework.
For automated RAG evaluation without gold labels, RAGAS \citep{es2024ragas} introduced
reference-free metrics for faithfulness and answer relevance, while ARES
\citep{saadfalcon2024ares} fine-tunes lightweight LLM judges with prediction-powered
inference.
\citet{li2025ragbestpractices} provide a systematic empirical study of RAG design
choices---chunk size, query expansion, retrieval stride---at COLING~2025.
Our work differs from prior benchmarks in two key respects: (i) we focus specifically
on \emph{retrieval method comparison} rather than LLM evaluation, and (ii) we target
documents with mixed text-and-table content, a setting underrepresented in existing
benchmarks. While T\textsuperscript{2}-RAGBench \citep{strich2026t2ragbench} evaluated
six retrieval methods with two metrics, we extend coverage to 15+ methods with
comprehensive retrieval and generation metrics, filling a critical gap in understanding
which retrieval strategies best serve heterogeneous financial documents.

% ============================================================
\section{Methodology}
\label{sec:method}

\subsection{Dataset: T\textsuperscript{2}-RAGBench}
\label{sec:dataset}

T\textsuperscript{2}-RAGBench \cite{strich2026t2ragbench} contains 23,088
question-context-answer triples derived from three financial QA datasets:
FinQA (8,281 pairs), ConvFinQA (3,458 pairs), and TAT-DQA (11,349 pairs).
The corpus comprises 7,318 unique documents with an average length of 920 tokens,
containing both narrative text and markdown-formatted tables.

All answers are numerical, and questions have been reformulated to be
context-independent (91.3\% validated by human experts).

\subsection{Retrieval Methods Under Evaluation}
\label{sec:methods}

% TODO: 1 paragraph per method

\subsection{Evaluation Metrics}
\label{sec:metrics}

\paragraph{Retrieval Metrics.}
We report Recall@$k$ ($k \in \{1, 3, 5, 10, 20\}$), Mean Reciprocal Rank (MRR@$k$),
normalized Discounted Cumulative Gain (nDCG@$k$), and Mean Average Precision (MAP).

\paragraph{Generation Metrics.}
Following T\textsuperscript{2}-RAGBench, our primary metric is Number Match (NM)
with relative tolerance $\epsilon = 10^{-2}$. We additionally report token-level F1,
ROUGE-L, and BERTScore.

\paragraph{Statistical Testing.}
We use paired bootstrap tests ($B = 10{,}000$) with Bonferroni correction
for multiple comparisons, reporting significance at $p < 0.05$.

\subsection{Experimental Setup}
\label{sec:setup}

% TODO: Hardware, models, hyperparameters, reproducibility details

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Main Retrieval Results}
\label{sec:main_results}

Table~\ref{tab:main_results} presents the retrieval performance of all evaluated methods
on the full T\textsuperscript{2}-RAGBench test set (23,088 queries over 7,318 documents).

\begin{table*}[t]
\centering
\small
\begin{tabular}{lcccccccc}
\toprule
\textbf{Method} & \textbf{R@1} & \textbf{R@3} & \textbf{R@5} & \textbf{R@10} & \textbf{R@20} & \textbf{MRR@3} & \textbf{nDCG@10} & \textbf{MAP} \\
\midrule
BM25                     & 0.293 & 0.552 & 0.644 & 0.735 & 0.797 & 0.411 & 0.515 & 0.449 \\
Dense (text-embed-3-large) & 0.248 & 0.481 & 0.587 & 0.703 & 0.798 & 0.351 & 0.466 & 0.398 \\
Hybrid (BM25+Dense, RRF)   & \textbf{0.308} & \textbf{0.588} & \textbf{0.695} & \textbf{0.801} & \textbf{0.877} & \textbf{0.433} & \textbf{0.551} & \textbf{0.477} \\
% Hybrid + Cohere Rerank  & --- & --- & --- & --- & --- & --- & --- & --- \\  % TODO: add when ready
% HyDE                    & --- & --- & --- & --- & --- & --- & --- & --- \\  % TODO
% Multi-Query + RRF       & --- & --- & --- & --- & --- & --- & --- & --- \\  % TODO
\bottomrule
\end{tabular}
\caption{Main retrieval results on T\textsuperscript{2}-RAGBench (23,088 queries).
Hybrid RRF consistently outperforms both sparse and dense retrieval across all metrics.
Best results per column in \textbf{bold}. Additional methods will be added as experiments complete.}
\label{tab:main_results}
\end{table*}

Hybrid retrieval (BM25 + Dense with Reciprocal Rank Fusion) achieves the best performance
across all metrics, with Recall@5 of 0.695 compared to 0.644 for BM25 and 0.587 for
dense retrieval alone---a relative improvement of 7.9\% and 18.4\%, respectively.

A notable finding is that BM25 outperforms dense retrieval (text-embedding-3-large)
on all metrics except Recall@20, where they are nearly tied (0.797 vs.\ 0.798).
This suggests that lexical matching is particularly effective for financial documents,
where precise terminology (company names, metric labels, fiscal periods) provides
strong retrieval signals that semantic embeddings may dilute.

\paragraph{Per-Subset Analysis.}
Table~\ref{tab:subset_results} breaks down performance by dataset subset.

\begin{table}[t]
\centering
\small
\begin{tabular}{llccc}
\toprule
\textbf{Subset} & \textbf{Method} & \textbf{R@5} & \textbf{R@10} & \textbf{MRR@3} \\
\midrule
\multirow{3}{*}{FinQA}
  & BM25    & 0.729 & 0.834 & 0.389 \\
  & Dense   & 0.611 & 0.748 & 0.308 \\
  & Hybrid  & \textbf{0.737} & \textbf{0.856} & \textbf{0.389} \\
\midrule
\multirow{3}{*}{ConvFinQA}
  & BM25    & 0.696 & 0.781 & 0.500 \\
  & Dense   & 0.654 & 0.781 & 0.410 \\
  & Hybrid  & \textbf{0.755} & \textbf{0.851} & \textbf{0.519} \\
\midrule
\multirow{3}{*}{TAT-DQA}
  & BM25    & 0.566 & 0.649 & 0.400 \\
  & Dense   & 0.549 & 0.647 & 0.364 \\
  & Hybrid  & \textbf{0.647} & \textbf{0.746} & \textbf{0.438} \\
\bottomrule
\end{tabular}
\caption{Per-subset retrieval results. TAT-DQA is the most challenging subset
across all methods. Hybrid RRF provides the largest improvement on TAT-DQA
(+8.1pp Recall@5 over BM25).}
\label{tab:subset_results}
\end{table}

TAT-DQA emerges as the most challenging subset across all methods (Recall@5: 0.647
for the best method vs.\ 0.755 for ConvFinQA), likely due to its emphasis on
diverse numerical operations over complex table layouts.
Hybrid fusion provides the largest absolute improvement on TAT-DQA (+8.1 percentage
points Recall@5 over BM25), suggesting that combining lexical and semantic signals
is especially valuable for table-heavy questions.

\subsection{End-to-End Generation Results}
\label{sec:generation}

% TODO: Generation results for top methods

\subsection{Analysis by Question Type}
\label{sec:subset_analysis}

% TODO: Per-subset (FinQA, ConvFinQA, TAT-DQA) breakdown

\subsection{Ablation Studies}
\label{sec:ablations}

% TODO: Embedding model, chunk size, reranker, fusion method ablations

\subsection{Error Analysis}
\label{sec:errors}

% TODO: Failure mode categorization

% ============================================================
\section{Discussion}
\label{sec:discussion}

% TODO: Key findings, practical recommendations, cost-accuracy trade-offs

\paragraph{Limitations.}
% TODO: Honest limitations

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

% TODO: Summary and future work

% ============================================================
\section*{Acknowledgments}
We thank Christopher Mierbach and Radiate for generously providing Azure AI compute credits
that made the large-scale experiments in this work possible.

\bibliography{references}
\bibliographystyle{acl_natbib}

\appendix

\section{Hyperparameter Details}
\label{app:hyperparams}

\section{Full Results Tables}
\label{app:full_results}

\section{Prompt Templates}
\label{app:prompts}

\end{document}
